Type-Safe Modular Parsing

### 2016-12-22 paper/sections/Overview.tex
Choosing the parsing technique.

\huang{done} \bruno{Write a summary for this section}

\huang{done}\bruno{Fill in all the missing references}

@@

Unfortunately many parser combinators have important limitations for
modular parsing. In particular several parser combinators,
including the famous library Parsec~\cite{Leijen2001} library, require
programmers to manually do \textit{left-recursion elimination}, and
require significant amounts \textit{backtracking}. Both of these are
problematic in a modular setting.

@@

\paragraph{Backtracking} The need for backtracking is also problematic
in a modular setting. For example, suppose that we only have \inlinecode{import..from} statements before, now we want to extend the grammar to add an \inlinecode{import..as} case. The third line of grammar below is the new case.

@@

Because the \inlinecode{import..from} case shares a prefix with the new case \inlinecode{import..as}, when the former case fails, we must backtrack to the beginning. Take Parsec as an example, its choice
combinator \inlinecode{<|>} only tries the second alternative if the first fails
without any token consumption. An auxiliary function \inlinecode{try} is used for explicit backtracking.

@@

Given the full grammar, we can decide where to put \inlinecode{try} for backtracking. However, in a modular setting we should always consider the worst case, that all alternatives may share common prefixes with future cases. Then we need to backtrack for all the branches, which results worst case exponential time complexity.
In Parsec, we must add \inlinecode{try} everywhere:

\begin{lstlisting}[language=PlainCode]
parser = try parseImpFrom <|> try parseA <|> try parseB <|> ... <|> try parseImpAs
\end{lstlisting}

\huang{I've rewritten this paragraph}\bruno{The example is good, I think but the explanation is not.
You want to say that when you know
  the full grammar, you can figure out where ``try'' is
  needed. Without the full grammar you'd need assume the worst
  case. Make an effort to make your explanation cristal clear!
I think you want to miss the abstract explanation that you give first,
with the explanation about the concrete example.
Start with ``The need for backtracking is also problematic
in a modular setting. For example, suppose ...'' and synchronize
the abstract explanation and the explanation for the example.
}

\paragraph{Packrat Parsing}
Fortunately more advanced parsing techniques such as Packrat
parsing~\cite{Ford2002}, address the limitations of simple parser combinators
such as Parsec. Packrat parsers use
memoization to record the result of applying each parser at each
position of the input, so that repeated computation is eliminated, and
they support left-recursive grammars directly~\cite{warth2008}. All of these properties are very
suitable for modularity, thus we use Packrat parsers as the underlying
parsing technique. In Section \ref{sec:packratparsers}, we will give
an overview of the Packrat parsing library implemented in Scala,
which we use.

It is worth mentioning that the choice of parser combinators will not
affect the other parts of our library. One can choose other parser
combinators like Parsec, in cases that the performance and supporting
of left-recursion are not major concerns.

>>>>

\section{To be discussed in Introduction}

\haoyuan{I suggest this part to be moved to Introduction, and only discuss why Packrat is selected among
different parser combinators in Overview.}

@@

In the last section, we have argued that parser combinators are a suitable parsing technique for
our purpose, as they naturally build modular parsers for type-checking.
Unfortunately many parser combinators have important limitations.
In particular several parser combinators,
including the famous Parsec~\cite{Leijen2001} library, require
programmers to manually do \textit{left-recursion elimination} and \textit{longest match composition}, and
require significant amounts of \textit{backtracking}. All of them are
problematic in a modular setting.

@@

\paragraph{Longest Match Composition}
Some parsing libraries provide combinators for alternative composition with longest match,
while others do not. Now consider the non-left-recursive grammar:
\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <int> \alt <int> `+' <expr>
\end{grammar}
In Parsec, for instance, the parser
\begin{lstlisting}[language=PlainCode]
parser = parseInt <|> parseAdd
\end{lstlisting}
will only parse the input \inlinecode{"1 + 2"} to \inlinecode{"1"}, as \inlinecode{parseInt} successfully parses \inlinecode{"1"}
and terminates parsing. With traditional alternative
composition in recursive descent parsers, the parser in the front successfully parses a substring of the input, then it unexpectedly
finishes parsing and discards the rest, in spite that latter parsers might be able to parse the whole input.
In contrast, \inlinecode{parser = parseAdd <|> parseInt} works as expected, yet manual composition for longest match is quite
inconvenient, especially in a modular setting. When the grammar is extended with new rules, programmers are supposed to adjust
the order of parsers, which is tedious.

\paragraph{Backtracking} The need for backtracking is also problematic
in a modular setting. For example, suppose that we only have \inlinecode{import..from} statements before,
but now we want to extend the grammar with the \inlinecode{import..as} case, as shown in the third line of grammar below.

@@

Since the \inlinecode{import..from} case shares a prefix with the new case \inlinecode{import..as}, when the former case fails, we must backtrack to the beginning. Take Parsec as an example, its choice
combinator \inlinecode{<|>} only tries the second alternative if the first fails
without any token consumption. An auxiliary function \inlinecode{try} is used for explicit backtracking.

@@

Given the full grammar, we can decide where to put \inlinecode{try} for backtracking. However, with modular parsing we are unable to overlook the full grammar, hence the worst case should be considered, namely all alternatives may share common prefixes with future cases. In that case we need to backtrack for all the branches.
To avoid failures in the future, we have to add \inlinecode{try} everywhere:

\begin{lstlisting}[language=PlainCode]
parser = try parseImpFrom <|> try parseA <|> try parseB <|> ... <|> try parseImpAs
\end{lstlisting}
which results in the worst exponential time complexity.

%\huang{I've rewritten this paragraph}\bruno{The example is good, I think but the explanation is not.
%You want to say that when you know
%  the full grammar, you can figure out where ``try'' is
%  needed. Without the full grammar you'd need assume the worst
%  case. Make an effort to make your explanation cristal clear!
%I think you want to miss the abstract explanation that you give first,
%with the explanation about the concrete example.
%Start with ``The need for backtracking is also problematic
%in a modular setting. For example, suppose ...'' and synchronize
%the abstract explanation and the explanation for the example.
%}

\paragraph{Packrat Parsing}
Fortunately some more advanced parsing techniques such as Packrat
parsing~\cite{Ford2002}, address the limitations of simple parser combinators
such as Parsec. Packrat parsers use
memoization to record the result of applying each parser at each
position of the input, so that repeated computation is eliminated.
Moreover, theoretically the algorithm behind Packrat parsers
supports both direct and indirect left-recursion ~\cite{warth2008}.
The current version of the library is still
buggy with indirect left-recursion, but we believe that
they will fix it in the future, and direct left-recursion is
already practical to use for a large number of applications. All of these properties are very
suitable for modularity, thus we decided to use Packrat parsers as the underlying
parsing technique in \name.

It is worth mentioning that the choice of parser combinators will not
affect the other parts of our library. One can choose other parser
combinators like Parsec, in cases that the performance and supporting
of left-recursion are not major concerns. A different library can even build a new
\name with fancy features or higher efficiency.
